### 2021-03-19 ch/1-introduction.tex
Address the remaining feedback

The overall goal in all cases is to make the classifier invariant to $s$.
We can measure the invariance to $s$ directly by using \emph{fairness metrics}
on the predictions of the classifier.
In contrast to the accuracy metric,
these fairness metrics give a more complete picture of how invariant the classifier is.
This is especially true if evaluation is done on an imbalanced set
-- for example because a truly unbiased test set is not available.
In this case, accuracy can be highly misleading,
because a good performance in the majority class can hide a poor performance in a minority class.
Fairness metrics do not suffer from this problem,
because they specifically look at the results in the different subgroups.
However, we typically try to evaluate our models on a test set that is as unbiased as possible,
in order to have a meaningful accuracy.

The fairness definition with the clearest interpretation in the described setup of label bias and sampling bias is \acf{DP},
also called statistical parity or independence.
It demands that the predictions $\hat{y}$ be independent of the sensitive attribute \(s\).
So, for binary $s$ and $y$:
\begin{align}
  P(\hat{y}=1|s=0) &= P(\hat{y}=1|s=1)~.
  \label{eq:dp-def}
\end{align}
There are multiple \ac{DP} \emph{metrics} which track how close the predictions are to satisfying the equality,
among them the difference and the ratio of the terms on the two sides of the equation.

When we previously talked about unbiased datasets,
we did not specify exactly what this meant,
and that is because this can differ from task to task,
but one way to define it is as a \emph{balanced} dataset where all combinations of $s$ and $y$ occur at the same rate:
\begin{align}
  \label{eq:balanced-dataset}
  P(y=0,s=0)=P(y=0,s=1)=P(y=1,s=0)=\dots
\end{align}
In such a dataset, we have $y \perp s$, and thus,
perfect predictions ($\hat{y}=y$) on this dataset will satisfy $\hat{y} \perp s$ and hence \ac{DP}.
We can conclude that perfect accuracy on a balanced test set implies \acl{DP} (the opposite does not hold).
However, if a model's predictions satisfy \ac{DP} on a \emph{biased} dataset,
then they cannot be perfectly accurate with respect to that dataset's biased labels anymore,
which makes sense because the goal is to be accurate to the \emph{unbiased} dataset.
This leads to a fairness-accuracy trade-off on biased test sets.

The other two fairness definitions which are commonly used are \ac{EOpp} and \ac{EOdds}.
These do not require the prediction $\hat{y}$ to be independent of $s$,
but they do require that the model makes equally high-quality predictions for all values of $s$.
In particular, for \ac{EOpp}, the \acp{TPR} need to be the same for all demographic groups
(again for binary $s$ and $y$):
\begin{align}
  \label{eq:eopp-def}
  P(\hat{y}=1|y=1,s=0) = P(\hat{y}=1|y=1,s=1)~.
\end{align}
\ac{EOdds} has the same requirement, but also extends it to the \acp{TNR}:
\begin{align}
  P(\hat{y}=y'|y=y',s=0) = P(\hat{y}=y'|y=y',s=1)\quad\forall y'~.
  \label{eq:eodds-def}
\end{align}
Just like \ac{DP},
\ac{EOpp} and \ac{EOdds} can help us evaluate how much a classifier is affected by the bias in the training set.

@@

\section{Structure of document}%
\label{sec:thesis-structure}
In \chapref{ch:related-work}, I present related work,
concentrating on the area of algorithmic fairness and dataset bias.
\Rangechapref{ch:paper1}{ch:paper3} constitute the main contribution of this thesis.
Each of these chapters starts with a brief section which positions the chapter in the context of the this document.
This is followed by the contents of publications.
In the final chapter, \chapref{ch:conclusion}, I present conclusions from the main work and directions for future work.

>>>>

The overall goal in all cases is to make the classifier invariant to $s$.
We can measure the invariance to $s$ directly by computing \emph{fairness metrics}
for the predictions of the classifier.
In contrast to the accuracy metric,
these fairness metrics give a more complete picture of how invariant the classifier is.
This is especially true if evaluation is done on an imbalanced set
-- for example in the event that a truly unbiased test set is unavailable.
In such cases, accuracy can be highly misleading,
because good performance on the majority class can hide poor performance on a minority class.
Fairness metrics do not suffer from this problem
because they specifically look at the results in the different subgroups.
However, we typically try to evaluate our models on a test set that is as unbiased as possible,
in order for accuracy to be meaningful.

The fairness definition with the clearest interpretation in the described setup of label bias and sampling bias is \acfi{DP},
also called \emph{statistical parity} or \emph{independence}.
It demands that the predictions $\hat{y}$ be independent of the sensitive attribute \(s\).
So, for binary $s$ and $y$:
\begin{align}
  P(\hat{y}=1|s=0) &= P(\hat{y}=1|s=1)~.
  \label{eq:dp-def}
\end{align}
There are multiple \ac{DP} \emph{metrics} which track how close the predictions are to satisfying the equality,
among them, the difference and the ratio of the terms on the two sides of the equation.

When we previously talked about unbiased datasets,
we did not specify exactly what this meant,
and that is because the definition thereof can differ from task to task,
but one way to define it is as a \emph{balanced} dataset where all combinations of $s$ and $y$ occur at the same rate:
\begin{align}
  \label{eq:balanced-dataset}
  P(y=0,s=0)=P(y=0,s=1)=P(y=1,s=0)=\dots
\end{align}
In such a dataset, we have $y \perp s$, and thus,
perfect predictions ($\hat{y}=y$) on this dataset will satisfy $\hat{y} \perp s$ and hence \ac{DP}.
We can conclude that perfect accuracy on a balanced test set implies \acl{DP} (the opposite does not hold).
However, if a model's predictions satisfy \ac{DP} on a \emph{biased} dataset,
then they cannot be perfectly accurate with respect to that dataset's biased labels anymore,
which makes sense because the goal is to be accurate to the \emph{unbiased} dataset.
This leads to a fairness-accuracy trade-off on biased test sets.

The other two fairness definitions which are commonly used are \acfi{EOpp} and \acfi{EOdds}.
These do not require the prediction $\hat{y}$ to be independent of $s$,
but they do require that the model makes equally high-quality predictions for all values of $s$.
In particular, for \ac{EOpp}, the \acfip{TPR} need to be the same for all demographic groups
(again for binary $s$ and $y$):
\begin{align}
  \label{eq:eopp-def}
  P(\hat{y}=1|y=1,s=0) = P(\hat{y}=1|y=1,s=1)~.
\end{align}
\ac{EOdds} has the same requirement, but also extends it to the \acp{TNR}:
\begin{align}
  P(\hat{y}=y'|y=y',s=0) = P(\hat{y}=y'|y=y',s=1)\quad\forall y'~.
  \label{eq:eodds-def}
\end{align}
Just like \ac{DP},
\ac{EOpp} and \ac{EOdds} can help us evaluate how much a classifier is affected by the bias in the training set.

@@

\section{Structure of document}%
\label{sec:thesis-structure}
In \chapref{ch:related-work}, I present related work,
concentrating on the area of algorithmic fairness and dataset bias.
\Chapref{ch:content} summarises the main contributions of this thesis
and describes their relationship to one another.
\Rangechapref{ch:paper1}{ch:paper3} reproduce the publications listed in the previous section.
% constitute the main contribution of this thesis.
% Each of these chapters starts with a brief section which positions the chapter in the context of the this document.
% This is followed by the contents of publications.
In the final chapter, \chapref{ch:conclusion}, I present conclusions from the main work and directions for future work.

### 2021-03-19 ch/2-related-work.tex
Address the remaining feedback

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2012fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalisation that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\section{Two views of the dataset bias problem}\label{two-views-of-the-dataset-bias-problem}
As hinted to in the introduction,
there is a divide in the relevant literature in how the problem of dataset bias is approached.
On the one hand, there is the ``ground-truth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is in use.

@@

On the other hand, there is the ``definition-centric'' view that,
given some training data and not necessarily knowing anything about the deployment setting,
we define \emph{fairness criteria} which a fair classifier should satisfy.
Here, the primary goal is not to correct for the deviation of the training data from the true distribution,
but simply to ensure that the classifier is fair in some specific sense.
By default, a classifier would \emph{not} be fair in that specific way.
It is left open whether this unfairness originates in the training data or in the algorithm itself.
Evaluation is typically performed on a test set that has all the same biases as the training set,
and thus a fairness-accuracy trade-off arises.

@@

\subsection{Precursors}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data mining.
Their first fundamental conclusion
is that it is not sufficient to remove the protected attribute
(which should not be used as a basis for prediction) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features%
\footnote{This is similar to the legal view of direct and indirect discrimination
where direct discrimination when people with a specific protected characteristic are treated worse,
whereas indirect discrimination is a policy that does not make direct references to a protected characteristic
but disproportionally affects people with a specific characteristic.}.
% The latter case can be
% detected by checking the other rules in the dataset.

\subsection{Modification of labels for fairness enforcement}%
\label{modification-of-labels-for-fairness-enforcement}
Independently of \citet{pedreshi2008discrimination}, but with a similar intention,
\citet{kamiran2009classifying} considered discrimination in classification.
Their goal is to modify the dataset so that any classifier trained on it will be ``fair'',
which is measured with a fairness metric on the test set predictions.
The test set is drawn from the same distribution as the training set.
To demonstrate their method, they use the ``German credit dataset''~\citep{Dua:2017}
where the task is to classify people as a good or a bad credit risk.

The authors formulate the problem as it was later done in other works on fairness as well:
there is a set of features \(x\), a binary class label \(y\)
and a special \emph{sensitive attribute} \(s\) which determines the demographic group.
The sensitive attribute \(s\) is assumed to be binary in the derivation,
but does not have to be.
\(s=0\) refers to the group vulnerable to discrimination and \(s=1\) to all other individuals.
It is further assumed that one of the class labels is generally desirable;
for example, it might correspond to being accepted for a credit or being given bail.
This class label is being referred to as the positive label: \(y = 1\).

The authors then define a concrete measure of discrimination (where $\hat{y}$ refers to the prediction of a classifier):
\begin{align}
  \label{eq:disc}
  Disc := P(\hat{y}=1|s =1) - P(\hat{y}=1|s = 0)~.
\end{align}
% (The paper uses a different notation.
% The notation here reflects the notation in later papers.)
It is zero when the individuals in both groups have the same chance to get a positive prediction.
Thus, it measures the violation of \ac{DP}, where $Dics=0$ corresponds to satisfying \ac{DP}.

Their algorithm for de-biasing the dataset comprises two steps.
First, a well-calibrated classifier (Na\"ive Bayes in the paper) is trained on the data
\emph{without} the sensitive attribute.
This classifier is then used to determine a ranking for how likely a positive label is for a given individual.
The highest ranked individuals that have label \(y=0\)
and are potentially discriminated against \(s=0\) get ``promoted'' to \(y=1\)
and the lowest ranked individuals with label \(y=1\) and \(s=1\) get ``demoted'' to \(y=0\)
until the dataset is fair according to the \(Disc\) measure.
The idea being that the ranking ensures that the changes in the dataset
happen to the most appropriate candidates (close to the decision boundary);
the main goal being a prevention of a big drop in accuracy of the final classifier
on the test set whose distribution matches the training set.

@@

This work was extended in \citet{calders2009building},
where the same discrimination criterion is used as before.
They present a different technique based on giving training examples different weights
instead of re-labelling them.
Examples with \(y=1\) and \(s=0\) get higher weight than those with \(y=0\) and \(s=0\).
For \(s=1\), those with \(y=0\) get higher weight than those with \(y=1\).
Weighting means that when the training examples are sampled from the dataset,
those with higher weight are used more often.
The two approaches are compared on the Adult/Census Income dataset \citep{kohavi1996scaling}
where the task is to estimate if someone earns more than \$50K per year.
The sensitive attribute is \emph{sex} (``gender'' in some publications).
% In addition to Na\"ive Bayes,
% two classifiers based on nearest neighbours and a decision tree classifier are used.
% The comparison shows that changing the labels leads to lower dependency (discrimination)
% than reweighting the examples.
% However, reweighting the examples leads to lower dependency than the baseline.
% The difference in accuracy is minor.

\subsection{Fair classifiers}\label{fair-classifiers}
Another common approach is to put fairness \emph{constraints} on the classifier
instead of manipulating the dataset.
\citet{calders2010three} proposes three such approaches,
which potentially give more control over the prediction bias than changing the datasets.
All three approaches are based on Na\"ive Bayes
and again aim to enforce \ac{DP} in the predictions as defined before.
The first approach shifts the predicted probabilities in a post-processing step
after training the classifier normally.
To this end, the sensitive attribute is treated differently
than the other features in the Na\"ive Bayes model.
Instead of the class being the cause of the sensitive attribute,
the sensitive attribute is regarded as one cause for the class.
This is a significant change in perspective that enables a better intuition about the problem.
In this modified Na\"ive Bayes,
the conditional probability \(P(y|s)\) is then modified in the already trained model
until the discrimination score is below a given threshold.
Care is taken to not distort the predicted distribution too much with respect to the unfair model.
The end result is a model that is (ideally) unbiased
but still relies on the sensitive attribute for predictions.
This is unworkable if the sensitive attributes are not always available when making predictions.
Nevertheless, the algorithm can be considered an improvement over manipulating the dataset,
as the resulting bias can be controlled to a much finer degree.

@@

The third methods introduces a new latent variable, called \(L\) in the paper.
\(L\) is intended to be an unbiased target class label
that replaces the biased class label \(y\) from the training data.
\(L\) can only be estimated and is unbiased
in the sense that it is statistically independent from sensitive attribute: \(P(L|s=0) = P(L|s=1)\).
The observed class label \(y\) then only depends on the latent label and the sensitive attribute.
\(L\) is found by using expectation maximisation
to iteratively search for a value that maximises the likelihood of the dataset.
The search is restricted by enforcing that \(L\) must be equal to \(y\) except
when \(s=0\) and \(y=0\) or when \(s=1\) and \(y=1\)
because these are the two cases where we expect discrimination.
There is other prior knowledge that can be incorporated into the search.
% This method seems to have been the most sophisticated at that point in time.

@@

\subsection{Fairness based on similarity}%
\label{fairness-based-on-similarity}
Nearly all previously discussed papers express some dissatisfaction
with the \acf{DP} fairness definition.
\citet{luong2011k} use a very different fairness metric
based on the idea that similar people should be treated similarly.
The setup is still that the training set is biased,
only a different fairness definition is used to judge classifiers.
The authors first define a distance metric which defines a neighbourhood,
based on the Manhattan distance on the z-scores of the attributes.
The distance metric is supposed to be only make use of legally admissible attributes.
An individual is then considered to be unfairly treated
if it is classified differently than its neighbours.
More concretely, for any data point we can check
how many of the \(k\) nearest neighbours have the same class label as that data point.
If the percentage is under a certain threshold
then there was discrimination against the individual corresponding to that data point.
In order to create a fair classifier, for this definition of fairness,
the authors propose to pre-process the dataset,
similar to \citet{kamiran2009classifying} before,
by flipping the class labels of those data points where the class label is considered unfair.
The experiments in the paper show that this method is successful in reducing the discrimination
(according to the given criterion)
for a range of classifiers on the Adult/Census Income dataset.

\citet{dwork2012fairness} give a more in-depth analysis
of the idea of distance-based fairness and \ac{DP}.
They present explicit detailed criticism of the \ac{DP} metric,
mainly in the form of several scenarios in which \ac{DP} is maintained,
but the system treats a lot of individuals very unfairly.
In one scenario, the sensitive attribute carries important information
and removing it makes everyone worse off.
% In the second scenario, a malicious actor achieves \ac{DP}
% by promoting the weakest members of the disadvantaged group to receive a positive label
% and then argues that fairness is not a worthwhile goal
% because these individuals will not be able to succeed.
% The goal of the actor here is to undermine the whole goal of avoiding discrimination.
Another scenario considers the case of subset targeting,
where there is discrimination against subgroups
which are not covered by the Demographic Parity metric
that only considers the major demographic groups.
Defining fairness via group membership will always leave open the possibility
to discriminate against ever smaller subgroups down to individuals.

Based on these considerations, \citet{dwork2012fairness} then argue for
\emph{individual fairness} instead of \emph{group fairness}.
They propose a method based on a similarity metric,
but do not give a general recipe to construct such a metric;
instead stating it has to be constructed on a case-by-case basis.

@@

The proposed method is then as follows: a classifier is fair
if and only if the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimisation.
% This is similar to how \citet{kamishima2012fairness} used the fairness criterion as a regulariser.
% However, \citet{dwork2012fairness} intend to realise it as a strict constraint.
% Furthermore, they show that their criterion is compatible with Demographic Parity
% but does not guarantee it in general.
% When the two demographic groups are similar according to the individual similarity measure,
% then the Lipschitz condition will lead to a good approximation of Demographic Parity.
% If the demographic groups are very different,
% then the Lipschitz criterion is essentially incompatible with Demographic Parity.

>>>>

\chapter{Related work}\label{ch:related-work}
%
% =========================
%      old introduction
% =========================
% \section{Introduction}\label{introduction}
% Fairness in Machine Learning is a topic that has recently received increasing attention.
% Machine Learning has become very widespread in our society
% which brings some dangers with it.
% Decisions such as credit eligibility and hiring are increasingly done automatically.
% However, the algorithms that are used in these decisions can have strong biases
% that stem from the training data.
% Often, the algorithms are even more biased than the training data.
% \citet{kamishima2012fairness} give a great illustration for this
% (adapted from \citet{calders2010three}).
% The illustration is based on the Adult/Census Income dataset \citep{kohavi1996scaling}
% which contains records of individuals with thirteen attributes, among them age and gender.
% The task is to predict whether or not the individual earns more than \$50K per year.
% In the dataset, 32\% of the men have a positive label (\ie, they earn more than \$50K),
% but of the women only 11\% have a positive label.
% With this ratio of about 3:1, the dataset is already skewed.
% However, when training a \href{https://en.wikipedia.org/wiki/Naive_Bayes_classifier}{Na\"ive Bayes classifier} on the data,
% the ratio in the predictions is even more unfair:
% the amount of women who get a positive prediction is even lower than 11\%.
% If the Na\"ive Bayes classifier does not get access to the gender information,
% the predictions will be a bit fairer,
% but they are still more unfair than the original dataset labels.
%
% This phenomenon of predictions being more unfair than an already skewed dataset
% is due to the generalisation that classifiers have to make.
% It is easier to classify all but the most outstanding women as low income
% than to learn a complicated decision boundary.
% Furthermore, removing the gender information does not solve the problem
% because the gender can be inferred from other data.
%
% The following is a review of the literature on Fairness in Machine Learning.
% Several different approaches to and definition of Fairness are presented.
% The focus is on identifying the major trends in the field from the beginning.
\section{Two views of the dataset bias problem}\label{two-views-of-the-dataset-bias-problem}
As alluded to in the introduction,
there is a divide in the relevant literature in how the problem of dataset bias is approached.
On the one hand, there is the ``ground-truth-centric'' view -- expressed through most of the introduction --
that the training set is biased but the deployment setting is not.
By \emph{deployment setting} I am here referring to the real distribution
that will be encountered by the machine learning system once it is in use.

@@

On the other hand, there is the ``definition-centric'' view that,
given some training data and not necessarily knowing anything about the deployment setting,
we define \emph{fairness criteria} which a fair classifier should satisfy.
Here, the primary goal is not to correct for the deviation of the training data from the true distribution,
but simply to ensure that the classifier is fair in some specific sense.
By default, a classifier would \emph{not} be fair in that specific way.
It is left open as to whether this unfairness originates from the training data or from the algorithm itself.
Evaluation is typically performed on a test set that has all the same biases as the training set,
and thus gives rise to a fairness-accuracy trade-off.

@@

\subsection{Precursors}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data-mining.
Their first fundamental conclusion
is that it is not sufficient to remove the protected attribute
(which should not be used as a basis for prediction) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features%
\footnote{This is similar to the legal view of direct and indirect discrimination
where direct discrimination when people with a specific protected characteristic are treated worse,
whereas indirect discrimination is a policy that does not make direct references to a protected characteristic
but disproportionally affects people with a specific characteristic.}.
% The latter case can be
% detected by checking the other rules in the dataset.

\subsection{Modification of labels for fairness enforcement}%
\label{modification-of-labels-for-fairness-enforcement}
Independently of \citet{pedreshi2008discrimination}, but with a similar intention,
\citet{kamiran2009classifying} considered discrimination in classification tasks.
Their goal is to modify the dataset so that any classifier trained on it will be ``fair'';
a condition which is measured with a fairness metric on the test set predictions.
The test set is drawn from the same distribution as the training set.
To demonstrate their method, they use the ``German credit dataset''~\citep{Dua:2017}
where the task is to classify people as a good or a bad credit risk.

The authors formulate the problem as it was later done in other works on fairness as well:
there is a set of features \(x\), a binary class label \(y\)
and a special \emph{sensitive attribute} \(s\) which determines the demographic group.
The sensitive attribute \(s\) is assumed to be binary in the derivation,
but does not have to be.
\(s=0\) refers to the group vulnerable to discrimination and \(s=1\) to all other individuals.
It is further assumed that one of the class labels is generally desirable;
for example, it might correspond to being accepted for a loan or being given bail.
This class label is being referred to as the positive label: \(y = 1\).

The authors then define a concrete measure of discrimination (where $\hat{y}$ refers to the prediction of a classifier):
\begin{align}
  \label{eq:disc}
  Disc := P(\hat{y}=1|s =1) - P(\hat{y}=1|s = 0)~.
\end{align}
% (The paper uses a different notation.
% The notation here reflects the notation in later papers.)
It is zero when the individuals in both groups have the same chance to get a positive prediction.
Thus, it measures the violation of \ac{DP}, where $Disc=0$ corresponds to satisfying \ac{DP}.

Their algorithm for de-biasing the dataset comprises two steps.
First, a well-calibrated classifier (Na\"ive Bayes in the paper) is trained on the data
\emph{without} the sensitive attribute.
This classifier is then used to determine a ranking for how likely a positive label is for a given individual.
The highest ranked individuals that have label \(y=0\)
and are potentially discriminated against \(s=0\) get ``promoted'' to \(y=1\)
and the lowest ranked individuals with label \(y=1\) and \(s=1\) get ``demoted'' to \(y=0\)
until the dataset is fair according to the \(Disc\) measure.
The idea being that the ranking ensures that the changes in the dataset
happen to the most appropriate candidates (those closest to the decision boundary);
the main goal being a prevention of a big drop in accuracy of the final classifier
on the test set whose distribution matches the training set.

@@

This work was extended in \citet{calders2009building},
where the same discrimination criterion is used as before.
They present a different technique based on giving training examples different weights
instead of re-labelling them.
Examples with \(y=1\) and \(s=0\) get higher weight than those with \(y=0\) and \(s=0\).
For \(s=1\), those with \(y=0\) get higher weight than those with \(y=1\).
Weighting means that when the training examples are sampled from the dataset,
those with higher weight are used more often.
The two approaches are compared on the Adult/Census Income dataset \citep{kohavi1996scaling}
where the task is to estimate if someone earns more than \$50K per year;
the sensitive attribute being \emph{sex} (``gender'' in some publications).
% In addition to Na\"ive Bayes,
% two classifiers based on nearest neighbours and a decision tree classifier are used.
% The comparison shows that changing the labels leads to lower dependency (discrimination)
% than reweighting the examples.
% However, reweighting the examples leads to lower dependency than the baseline.
% The difference in accuracy is minor.

\subsection{Fair classifiers}\label{fair-classifiers}
Another common approach is to put fairness \emph{constraints} on the classifier
instead of manipulating the dataset.
\citet{calders2010three} proposes three such approaches,
which potentially give more control over the prediction bias than changing the datasets.
All three approaches are based on Na\"ive Bayes
and again aim to enforce \ac{DP} in the predictions, as defined before.
The first approach shifts the predicted probabilities in a post-processing step
after training the classifier normally.
To this end, the sensitive attribute is treated differently
than the other features in the Na\"ive Bayes model.
Instead of the class being the cause of the sensitive attribute,
the sensitive attribute is regarded as one cause for the class.
This is a significant change in perspective that enables a better intuition about the problem.
In this modified Na\"ive Bayes,
the conditional probability \(P(y|s)\) is then modified in the already trained model
until the discrimination score is below a given threshold.
Care is taken to not distort the predicted distribution too much with respect to the unfair model.
The end result is a model that is (ideally) unbiased
but still relies on the sensitive attribute for predictions.
This is unworkable if the sensitive attributes are not always available when making predictions.
Nevertheless, the algorithm can be considered an improvement over manipulating the dataset,
as the resulting bias can be controlled to a much finer degree.

@@

The third method introduces a new latent variable, called \(L\) in the paper.
\(L\) is intended to be an unbiased target class label
that replaces the biased class label \(y\) from the training data.
\(L\) can only be estimated and is unbiased
in the sense that it is statistically independent from sensitive attribute: \(P(L|s=0) = P(L|s=1)\).
The observed class label \(y\) then only depends on the latent label and the sensitive attribute.
\(L\) is found by using expectation maximisation
to iteratively search for a value that maximises the likelihood of the dataset.
The search is restricted by enforcing that \(L\) must be equal to \(y\) except
when \(s=0\) and \(y=0\) or when \(s=1\) and \(y=1\)
because these are the two cases where we expect discrimination.
There is other prior knowledge that can be incorporated into the search.
% This method seems to have been the most sophisticated at that point in time.

@@

\subsection{Fairness based on similarity}%
\label{fairness-based-on-similarity}
Nearly all previously discussed papers express some dissatisfaction
with the \acf{DP} fairness definition.
\citet{luong2011k} use a very different fairness metric
based on the idea that similar people should be treated similarly.
The setup is still that the training set is biased,
only a different fairness definition is used to judge classifiers.
The authors first define a distance metric which defines a neighbourhood,
based on the Manhattan distance on the z-scores of the attributes.
The distance metric is supposed to be only making use of legally admissible attributes.
An individual is then considered to be unfairly treated
if it is classified differently than its neighbours.
More concretely, for any data point we can check
how many of the \(k\) nearest neighbours have the same class label as that data point.
If the percentage is under a certain threshold
then there was discrimination against the individual corresponding to that data point.
In order to create a fair classifier, for this definition of fairness,
the authors propose to pre-process the dataset,
similar to \citet{kamiran2009classifying} before,
by flipping the class labels of those data points where the class label is considered unfair.
The experiments in the paper show that this method is successful in reducing the discrimination
(according to the given criterion)
for a range of classifiers on the Adult/Census Income dataset.

\citet{dwork2012fairness} give a more in-depth analysis
of the idea of distance-based fairness and \ac{DP}.
They present explicit detailed criticism of the \ac{DP} metric,
mainly in the form of several scenarios in which \ac{DP} is maintained,
but the system treats a lot of individuals very unfairly.
In one scenario, the sensitive attribute carries important information
and removing it makes everyone worse off.
% In the second scenario, a malicious actor achieves \ac{DP}
% by promoting the weakest members of the disadvantaged group to receive a positive label
% and then argues that fairness is not a worthwhile goal
% because these individuals will not be able to succeed.
% The goal of the actor here is to undermine the whole goal of avoiding discrimination.
Another scenario considers the case of \emph{subset targeting},
where there is discrimination against subgroups
which are not covered by the \ac{DP} metric
that only considers the major demographic groups.
Defining fairness via group membership will always leave open the possibility
of discriminating against ever smaller subgroups down to the level  of individuals.

Based on these considerations, \citet{dwork2012fairness} argue for
\emph{individual fairness} instead of \emph{group fairness}.
They propose a method based on a similarity metric,
but do not give a general recipe to construct such a metric;
instead stating it has to be constructed on a case-by-case basis.

@@

The proposed method is then as follows: a classifier is fair
\emph{if and only if} the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimisation.
% This is similar to how \citet{kamishima2012fairness} used the fairness criterion as a regulariser.
% However, \citet{dwork2012fairness} intend to realise it as a strict constraint.
% Furthermore, they show that their criterion is compatible with Demographic Parity
% but does not guarantee it in general.
% When the two demographic groups are similar according to the individual similarity measure,
% then the Lipschitz condition will lead to a good approximation of Demographic Parity.
% If the demographic groups are very different,
% then the Lipschitz criterion is essentially incompatible with Demographic Parity.
